%File: anonymous-submission.tex
%
% AAAI 2027 Anonymous Submission.
% Requires the AAAI author kit files in this directory:
%     aaai2027.sty   aaai2027.bst
% (download from the same AAAI author-kit package as the template PDFs).
%
% Compile with:  pdflatex -> bibtex -> pdflatex -> pdflatex

\documentclass[letterpaper]{article}
\usepackage{aaai2027}  % AAAI Press style; anonymity comes from the author field

\usepackage{times}      % roman font
\usepackage{helvet}     % sans-serif font
\usepackage{courier}    % typewriter font
\usepackage[hyphens]{url}
\usepackage{graphicx}
\urlstyle{rm}
\def\UrlFont{\rm}
\usepackage{natbib}
\usepackage{caption}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{algorithm}
\usepackage{algpseudocode}
\frenchspacing
\setlength{\pdfpagewidth}{8.5in}
\setlength{\pdfpageheight}{11in}

% AAAI required PDF metadata
\pdfinfo{
/TemplateVersion (2027.1)
}

% ---------------------------------------------------------------------------
\title{Learning to Branch and Cut by Imagining:\\ A World Model for Branch-and-Cut Search}
\author{Anonymous Submission}
\affiliations{Anonymous Affiliation}
% ---------------------------------------------------------------------------

\begin{document}
\maketitle

\begin{abstract}
Modern solvers for Mixed-Integer Programming rely on Branch-and-Cut, which
interleaves two heuristic decisions at every node of the search tree: which
cutting planes to add in order to tighten the linear-programming (LP)
relaxation, and which fractional variable to branch on. Both decisions are
costly to make well---strong branching, for example, evaluates every candidate
by solving its child LPs---and recent learning-based methods have addressed
each of them in isolation, scoring candidates only from the state of the current
node. We introduce a \emph{world model} for Branch-and-Cut: a learned latent
dynamics model that lets the solver \emph{imagine} the downstream consequences
of a decision rather than computing them with LP solves. A bipartite graph
encoder maps each node to a latent state, and a causal Transformer predicts how
that state, together with the embeddings of the individual variables within it,
evolves after an action. This makes it possible to roll a decision several steps
forward in latent space and evaluate the branching sequence it induces. Each
candidate is scored by a MuZero-style return that combines a learned per-step
reward with a cost-to-go value estimating the remaining size of the tree. The
same world model also selects globally valid cutting planes that it propagates
across the subtree, chooses branching variables, and orders the open nodes, all
while leaving the solver's optimality guarantee intact. We evaluate on Set Cover
across difficulty tiers, measuring the reduction in explored nodes relative to a
strong solver baseline and ablating each component of the planner.
\end{abstract}

% ---------------------------------------------------------------------------
\section{Introduction}

Mixed-Integer Programming (MIP) is a workhorse for combinatorial decision
making, and modern solvers attack it with Branch-and-Cut. Starting from the LP
relaxation of a problem, the solver repeatedly does two things at each node of
a search tree: it adds \emph{cutting planes}---valid inequalities that tighten
the relaxation without removing any integer solution---and, when the relaxation
remains fractional, it \emph{branches}, splitting the problem on a fractional
variable. The number of nodes the solver must explore before it can prove
optimality is governed almost entirely by how well these two decisions are
made.

Unfortunately, making them well is expensive. The gold standard for branching,
\emph{strong branching}, tentatively solves the child LPs of every candidate
variable and keeps the one that improves the bound the most; cut selection
faces an analogous trade-off between the many valid inequalities that could be
added at a node. A now-common response is to \emph{learn} these decisions:
graph neural networks trained to imitate strong branching amortize its cost into
a single cheap forward pass \citep{gasse2019exact,khalil2016learning}, and
reinforcement learning has been used to select cutting planes
\citep{tang2020reinforcement}. These methods are effective, but they share a
structural limitation. Each scores a candidate \emph{from the current node
alone}, discarding the very signal that makes strong branching powerful---the
quality of the subtree that a decision ultimately produces---and replacing it
with a one-shot ranking.

We argue that this gap is naturally closed by a \emph{world model}: a learned
model of the search dynamics that lets the solver \emph{imagine} the
consequences of a decision instead of paying for them with LP solves. We encode
each node of the search as a latent state and train a causal Transformer to
predict how that state, and the embeddings of the individual variables it
contains, evolve after an action is taken. Because the model predicts
per-variable embeddings and not only a summary vector, the branching policy can
be re-applied to an \emph{imagined} state, allowing us to unroll a whole
branching sequence in latent space and judge a candidate by the future it leads
to. Candidates are scored by a MuZero-style return that adds a learned per-step
reward along the rollout to a cost-to-go value estimating how many nodes remain.
The same latent model drives the solver's other discrete choices: it scores
candidate cutting planes against the node embedding and selects the ones worth
adding, and it orders the frontier of open nodes by predicted remaining work.
Throughout, the model influences only which decisions the solver makes and in
what order, never the bounds or feasibility tests themselves, so exactness is
preserved. This is the same principle behind model-based planning agents such as
MuZero and Dreamer \citep{schrittwieser2020mastering,hafner2020dream}, here
brought to the interior of an optimization solver.

Our contributions are as follows. We introduce a latent world model that plans
Branch-and-Cut decisions by simulating their consequences in latent space
rather than through LP solves. We show that a single such model can drive
branching, node selection, and the selection of globally valid cutting planes
that are propagated across the subtree, while preserving optimality. We propose
a cost-to-go value that targets the solver's true objective---the number of
nodes explored---and that can be trained directly from collected traces without
requiring them to be depth-first ordered. We stabilize the multi-step rollout
with a grounded, MuZero-style objective that combines a per-step reward, a value
bootstrap, and latent overshooting. Finally, we present an ablation that isolates
the contribution of each planning component to the reduction in explored nodes.

% ---------------------------------------------------------------------------
\section{Related Work}

Learning to make solver decisions is an active area \citep{bengio2021machine}.
For branching, \citet{khalil2016learning} learn scores from strong-branching
labels, and \citet{gasse2019exact} represent a node as a bipartite
constraint--variable graph and train a graph convolutional network to imitate
strong branching; the latter architecture is the backbone we build on.
\citet{nair2020solving} scale neural MIP solving with learned diving and
branching, and \citet{scavuzzo2022learning} train a branching policy with
reinforcement learning using tree-structured returns. For cutting planes,
\citet{tang2020reinforcement} cast cut selection as a sequential decision
problem and learn it with reinforcement learning, and learned large neighborhood
search \citep{sonnerat2021learning} targets yet another solver decision. All of
these methods evaluate a candidate from the state of the current node.

Our work is closer in spirit to model-based reinforcement learning. MuZero
\citep{schrittwieser2020mastering} plans with a learned dynamics model that is
never grounded in reconstructed observations, while Dreamer and its predecessors
\citep{hafner2020dream} learn behaviors by imagination in a latent space, using
latent overshooting to keep multi-step predictions faithful. We bring this style
of latent-space planning inside a Branch-and-Cut solver: rather than ranking
candidates by a static score, we simulate the subtree a decision induces and
select the candidate whose imagined future is best, and we do so for cutting
planes and node selection as well as for branching.

% ---------------------------------------------------------------------------
\section{Method}

We view Branch-and-Cut as a sequential decision process. At each node $t$ of the
search the solver observes a state $s_t$ (the current relaxation), takes an action
$a_t$ (branch on a fractional variable, or add a cut), and transitions to the next
node. We learn (i) an encoder that maps $s_t$ to a latent state, (ii) a dynamics
model that predicts how that latent evolves under an action, and (iii) heads that
score decisions by rolling the dynamics forward. Throughout, $d$ denotes the
embedding dimension.

\subsection{Node encoder}
Each node is a bipartite graph over the constraints and variables of the
relaxation: variable nodes carry LP-derived features (solution value, reduced
cost, bound and basis status), constraint nodes carry their features, and each
edge carries its coefficient. A graph network with GATv2 attention
\citep{brody2022how} performs two-directional message passing, and a learned
attention readout pools the result:
\begin{equation}
\big(\{h_i\}_{i=1}^{n},\, z\big) = \mathrm{Enc}(s_t),
\qquad h_i, z \in \mathbb{R}^{d},
\end{equation}
where $h_i$ is the embedding of variable $i$ and $z$ is a graph-level summary.
The embedding of an action that branches on variable $v$ is $a = h_v$.

\subsection{Latent dynamics}
A causal Transformer models the evolution of the latent along a trajectory. From
the sequence of tokens $u_\tau = W_{\mathrm{in}}[z_\tau \,\|\, a_\tau]$
($\tau \le t$) with learned positional encodings, it predicts the next graph
latent, while a per-variable head predicts how each variable embedding changes:
\begin{align}
z_{t+1} &= W_{\mathrm{out}}\,\mathrm{TF}(u_{0:t}), \\
h_i^{\,t+1} &= \mathrm{LN}\!\big(h_i^{\,t} + \mathrm{MLP}([\,h_i^{\,t}\,\|\,z_{t+1}\,\|\,a_t\,])\big).
\label{eq:varupdate}
\end{align}
Predicting per-variable embeddings~\eqref{eq:varupdate}, rather than only the
graph summary $z$, is what lets the branching policy be re-applied to an
\emph{imagined} state during planning. We ground the latent with a linear head
$\hat b_{t+1}=w_b^\top z_{t+1}$ that predicts the node's normalized dual bound,
tying it to a real solver quantity, and a reward head
$\hat r_{t}=w_r^\top z_{t+1}$ that predicts the per-step dual-bound improvement.

\subsection{Value, cost-to-go, and policy}
Let $\bar h_{\mathrm{frac}}$ be the mean embedding of the currently fractional
variables. The value and cost-to-go heads share this enriched input:
\begin{align}
V(z) &= \mathrm{MLP}_V([z \,\|\, \bar h_{\mathrm{frac}}]), \\
C(z) &= \mathrm{softplus}\big(\mathrm{MLP}_C([z \,\|\, \bar h_{\mathrm{frac}}])\big),
\end{align}
where $C$ predicts $\log(1+\text{remaining nodes})$. Its regression target is the
Monte-Carlo return $T-t$ read directly off a length-$T$ trajectory---the number of
steps still to come---which requires \emph{no} assumption that the trajectory is
depth-first ordered, unlike a subtree-size target. The branching policy is a
pointer network scoring each variable against the graph latent,
$\pi(v)\!\propto\! w^\top\!\tanh(W_k h_v + W_z z)$.

\subsection{Planning rollout for candidate selection}
To branch, we take the policy's top-$k$ candidates and score each by imagining the
subtree it induces. Starting from the current $(z,\{h_i\})$ with first action
$a=h_c$, we unroll the dynamics for $K$ steps, re-applying the policy to each
predicted state to choose the next action
$v_j=\arg\max_{v\in\mathcal{F}}\pi(v\mid z_j,\{h^j\})$ over the fractional set
$\mathcal{F}$. The candidate is scored by a MuZero-style return that accumulates
predicted rewards along the imagined path, bootstraps with the value at the leaf,
and penalizes predicted remaining work:
\begin{equation}
R(c) = \sum_{j=1}^{K}\gamma^{\,j-1}\hat r_j \;+\; \gamma^{\,K-1}V(z_K)
       \;-\; w_c\sum_{j=1}^{K}\gamma^{\,j-1}\hat C(z_j).
\label{eq:return}
\end{equation}
We expand a shallow tree of branching factor $b$ (the top-$b$ actions at each
step, child returns averaged); $b\!=\!1$ recovers a single greedy rollout. We
branch on $c^\star=\arg\max_c R(c)$. The same value drives \emph{node selection}:
open nodes are ordered by ascending predicted cost-to-go $C(z)$, so the subtree
expected to close soonest is explored first. Because these choices affect only
search order, never the bounds or feasibility tests, exactness is preserved.
Algorithm~\ref{alg:select} summarizes candidate selection.

\begin{algorithm}[t]
\caption{Branching-variable selection by latent rollout}
\label{alg:select}
\begin{algorithmic}[1]
\State \textbf{Input:} node latent $z$, variable embeddings $\{h_i\}$,
fractional set $\mathcal{F}$, depth $K$, factor $b$
\State $\mathcal{C}\gets \textsc{TopK}_k\big(\pi(\cdot\mid z,\{h_i\}),\,\mathcal{F}\big)$
\For{each candidate $c\in\mathcal{C}$}
  \State $R(c)\gets \textsc{Rollout}(z,\{h_i\},c,K,b)$ \Comment{Eq.~\eqref{eq:return}}
\EndFor
\State \textbf{return} $\arg\max_{c\in\mathcal{C}} R(c)$
\end{algorithmic}
\end{algorithm}

\subsection{Cutting-plane selection}
At each eligible node we generate globally valid cuts. For two tight covering
constraints $i_1,i_2$ (rows of $A$) with shared support
$S=\{j : A_{i_1 j}\!=\!A_{i_2 j}\!=\!1\}$, the intersection cut
$\sum_{j\in S} x_j \ge 1$ is a Chv\'atal--Gomory inequality with multipliers
$\tfrac{1}{2}(e_{i_1}+e_{i_2})$ rounded up. Such cuts hold for \emph{every}
integer-feasible solution regardless of the branching above, so they are safely
inherited by all descendants of the node that generates them---genuine
Branch-and-Cut rather than cut-and-branch. A pointer head scores each candidate
cut $\ell$ (with feature vector $\phi_\ell$: violation, efficacy, density,
parallelism, objective cutoff, support fraction) against the node latent,
$\rho_\ell = w^\top\tanh(W_k\,\mathrm{MLP}(\phi_\ell)+W_z z)$, and cuts with
$\sigma(\rho_\ell)>\tau$ are added, replacing fixed heuristics such as maximum
violation.

\subsection{Training}
The components are trained in a five-stage curriculum, with the objectives:
\begin{itemize}
  \item \textbf{Policy} (imitation): masked cross-entropy of $\pi$ against the
  strong-branching choice over the candidate set.
  \item \textbf{Value}: Huber loss between $V(z)$ and the normalized dual bound.
  \item \textbf{Dynamics}: an MSE\,$+$\,cosine latent-prediction loss on
  $z_{t+1}$; a reconstruction loss on the predicted $h_i^{\,t+1}$; grounding and
  reward losses on $\hat b,\hat r$; and \emph{latent overshooting}---unrolling
  $K$ steps from the model's own predictions and supervising each against the
  real future latent---to reduce compounding error at rollout time.
  \item \textbf{Joint fine-tuning}: the above plus the cost-to-go loss (Huber on
  $\log(1{+}(T{-}t))$) and a value-consistency term $\lVert V(\tilde z)-V(z)\rVert^2$
  that trains the value head to read dynamics-\emph{predicted} latents
  $\tilde z$ in distribution.
  \item \textbf{Cut selection}: weighted binary cross-entropy of $\rho_\ell$
  against a cut-quality label.
\end{itemize}
Class imbalance in the leaf-detection and cut-selection labels is corrected with a
positive-class weight equal to the negative/positive ratio, and early stopping on
a validation metric selects the best checkpoint in each stage.

% ---------------------------------------------------------------------------
\section{Experiments}

\subsection{Setup}
We evaluate on Set Cover instances generated with Ecole
\citep{prouvost2020ecole} across easy, medium, and hard difficulty tiers, and we
collect strong-branching trajectories for training. Our primary baseline is the
SCIP solver with its default pseudocost branching; we also compare against random
branching. Because the size of the search tree is what determines a solver's
cost, our primary metric is the number of explored nodes, with wall-clock time
reported as a secondary measure.

\subsection{Ablation}
To understand where any improvement comes from, we ablate the planner one
component at a time, starting from the imitation policy alone and successively
enabling the value-based rollout, the cost-to-go objective, the tree-structured
rollout, the reward-based return, learned node selection, and learned cut
selection.

\subsection{Results}
% [TODO: insert results once training completes.]
Table~\ref{tab:main} reports the reduction in explored nodes relative to SCIP
pseudocost branching. Following standard reproducibility guidance, we report
results as a mean with standard deviation over a held-out instance set and assess
each improvement with a Wilcoxon signed-rank test on the paired per-instance node
counts.

\begin{table}[t]
\centering
\begin{tabular}{lccc}
\toprule
Method & Nodes & vs.\ SCIP & $p$ \\
\midrule
SCIP pseudocost      & -- & -- & -- \\
Learned policy only  & -- & -- & -- \\
\ + value rollout    & -- & -- & -- \\
\ + cost-to-go       & -- & -- & -- \\
\ + tree rollout     & -- & -- & -- \\
\ + learned cuts     & -- & -- & -- \\
\bottomrule
\end{tabular}
\caption{Reduction in explored nodes and paired significance. [Placeholder;
filled once training completes.]}
\label{tab:main}
\end{table}

\subsection{Reproducibility}
Experiments were run on NVIDIA Quadro RTX 5000 (16\,GB) GPUs under CUDA~13.2,
using PyTorch and PyTorch Geometric. All seeds, hyperparameters, and source code
are provided.

% ---------------------------------------------------------------------------
\section{Conclusion}
We have presented a world model for Branch-and-Cut that plans the solver's
discrete decisions---branching, node selection, and the selection of globally
valid cutting planes---by imagining their consequences in latent space rather
than computing them with LP solves, all while preserving the solver's optimality
guarantee. Planning inside the search tree is only one way a world model can
serve optimization; extending the same idea to improvement search, such as
learned large neighborhood search, is a natural next step.

% ---------------------------------------------------------------------------
\bibliographystyle{aaai2027}
\bibliography{references}

\end{document}
